---
fontsize: 11pt
geometry: margin=1in
classoption:
 - landscape
linestretch: 1.25 # adjusts line spacing using the setspace package, e.g. 1.25, 1.5
indent: false # indent paragraphs
papersize: a4 # paper size, e.g. letter, a4
title: Quick demo 
output:
  pdf_document
pdf_engine: xelatex
---

The following *vignette* will provide some useful information about a real use case of the library, occupiying the data from the Chilean Presidential Elections.

# Data loading

Let's note that the package already includes the dataset that we're interested in, hence, it'll only be neccesary to load it in. 

```{r}
library(fastei)

df <- get("chile_election_2021")
head(df)
```

Within the essence of working with elections data, it's important to reduce our observations among electoral districts. Hence, we'll define a function to filter by `ELECTORAL.DISTRICT` and return the results as an `eim` object

```{r}
# function that filters a specific electoral district from the Chilean election dataset
get_XW_Chile <- function(elect_district, merge_blank_null = TRUE, remove_mismatch = TRUE) {
    #  read dataset
    df <- get("chile_election_2021")
    # filter a specific electoral district
    df_ed <- df[df$ELECTORAL.DISTRICT == elect_district, ]
    # remove ballot boxes with mismatch of votes and voters
    if (remove_mismatch) {
        df_ed <- df_ed[df_ed$MISMATCH == FALSE, ]
    }
    # get columns of candidates' votes
    X <- df_ed[df_ed$ELECTORAL.DISTRICT == elect_district, c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "BLANK.VOTES", "NULL.VOTES")]
    # merge blank and null votes
    if (merge_blank_null) {
        X$C8 <- X$BLANK.VOTES + X$NULL.VOTES
    }
    X <- X[, c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")]
    # convert X to matrix
    X <- as.matrix(X[, c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")])

    # get voters of ecah demographic group, and convert to matrix
    W <- as.matrix(df_ed[df_ed$ELECTORAL.DISTRICT == elect_district, c("X18.19", "X20.29", "X30.39", "X40.49", "X50.59", "X60.69", "X70.79", "X80.")])

    obj <- eim(X = X, W = W)
    return(obj)
}
```

# Setting the algorithm

Before running directly the algorithm there are some things that can be done. First of all, we would like to know what's the best group aggregation such that the log-likelihood is maximized. There's usually a reward for aggregating groups within big districts.

